Constructor specialisation in GHC

### 2013-07-19 2013specconstr/Main.tex
ok

\begin{introduction}
In Haskell, loops, even those of higher-order combinators such as @map@, are expressed as recursive functions. 
With boxed objects such as @Int@, each iteration ends up allocating new objects in memory, only to be deconstructed immediately in the next iteration.
For high-performance code, these spurious allocations turn out to have serious penalties.

This problem has been solved specifically for boxed objects such as @Int@, @Float@ and others in the worker/wrapper transform~\CITE.
However, in the case of code produced by stream fusion~\CITE, other types such as @Either@ and tuples are used as loop state.
Constructor specialisation, or SpecConstr, is a generalisation of worker/wrapper for arbitrary types.
This is able to remove such allocations in stream-fusion code, and in some circumstances, produce object code that competes with hand-written C.
\end{introduction}

\section{}

>>>>

\begin{introduction}
In Haskell, loops are expressed as recursive functions. 
When loop variables are updated, each iteration ends up allocating new objects in memory, only to be deconstructed immediately in the next iteration.
For high-performance code, these spurious allocations turn out to have serious penalties.

This problem has been solved specifically for boxed objects such as @Int@, @Float@ and others in the worker/wrapper transform~\CITE.
However, in the case of code produced by stream fusion~\CITE, other types such as @Either@ and tuples are used as loop variables.
Constructor specialisation, or SpecConstr, is a generalisation of worker/wrapper for arbitrary types.
This is able to remove such allocations in stream-fusion code, and in some circumstances, produce object code that competes with hand-written C.
\end{introduction}

\section{Background}
\subsection{Core}
Core is GHC's intermediate language, an explicitly typed System-F style language.
After the source program has been parsed, its types checked and inferred, it is converted to core.
The majority of optimisations are then performed on the core code,
before eventually being converted to object code.

To an experienced Haskell programmer, learning to read core should take little effort, and is a valuable skill for writing high-performance code.
Where clauses are turned into @let@s, or @letrec@s for recursive expressions.
Pattern matching and guards are converted to explicit @case@ expressions.

To illustrate this, there follows an example implementation of @foldl@, and its core equivalent.
\begin{code}
foldl :: (a -> b -> a) -> a -> [b] -> a
foldl k z xs = go xs z
 where
  go []     acc = acc
  go (x:xs) acc = go xs (k acc x)
\end{code}

The core is quite a bit noisier.

\begin{code}
foldl = \@a @b k z xs ->
 letrec {
     go = \xs' acc ->
       case xs' of {
         []     -> acc;
         : x xs -> go xs (k acc x)
       }
 } in go xs z
\end{code}

\subsection{Unboxed types}
In GHC, the @Int@ type is defined as a wrapper around an unboxed int,
and unboxed int literals and operators have @#@ appended to their name.

\section{Motivation}
Consider a simple example of summing a list of @Int@s.
\begin{code}
sum :: [Int] -> Int
sum = foldl (+) 0
\end{code}

After inlining @foldl@, we get a recursive worker function.
\begin{code}
data Int = I# Int#

sum xs = go xs (I# 0#)
 where
  go []          (I# acc) = I# acc
  go ((I# x):xs) (I# acc) = go xs (I# (acc +# x))
\end{code}

\begin{code}
dotp :: Vector Double -> Vector Double -> Double
dotp as bs = go (Nothing, 0) 0
 where
  go (_, i) acc
   | i > V.length as
   = acc
  go (Nothing, i) acc
   = go (Just (as!i), i) acc
  go (Just a, i) acc
   = go (Nothing, i + 1) (acc + (a * bs!i))
\end{code}

\section{Constructor specialisation}
